---
title: '6\. Worksheet: Among Site (Beta) Diversity -- Part 2'
author: "Student Name; Z620: Quantitative Biodiversity, Indiana University"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: pdf_document
geometry: margin=2.54cm
---

## OVERVIEW

In this worksheet, we continue to explore concepts, statistics, and visualizations related to $\beta$-diversity.
Now that you know how to formally quantify $\beta$-diversity, we will learn how to test hypotheses about $\beta$-diversity using multivariate statistics. 

## Directions:
1. In the Markdown version of this document in your cloned repo, change "Student Name" on line 3 (above) with your name.
2. Complete as much of the worksheet as possible during class.
3. Use the handout as a guide; it contains a more complete description of data sets along with examples of proper scripting needed to carry out the exercises.
4. Answer questions in the worksheet.
Space for your answers is provided in this document and is indicated by the ">" character.
If you need a second paragraph be sure to start the first line with ">".
You should notice that the answer is highlighted in green by RStudio (color may vary if you changed the editor theme). 
5. Before you leave the classroom today, you should **push** this file to your GitHub repo, at whatever stage you are. This will enable you to pull your work onto your own computer.
6. When you have completed the worksheet, **Knit** the text and code into a single PDF file by pressing the `Knit` button in the RStudio scripting panel.
This will save the PDF output in your Posit.cloud workspace: `/cloud/project/QB-2025/Week4-Beta/`
7. After Knitting, please submit the worksheet by making a **push** to your GitHub repo and then create a **pull request** via GitHub.
Your pull request should include this file (**6.BetaDiversity_2_Worksheet.Rmd**) 
with all code blocks filled out and questions answered) and the PDF output of `Knitr`  
(**6.BetaDiversity_2_Worksheet.pdf**).

The completed exercise is due on **Wednesday, February 12^th^, 2025 before 12:00 PM (noon)**.

## 1) R SETUP

Typically, the first thing you will do in either an R script or an RMarkdown file is setup your environment. 
This includes things such as setting the working directory and loading any packages that you will need.

In the R code chunk below, provide the code to:  

1. clear your R environment,
2. print your current working directory,
3. set your working directory to your `Week4-Beta/` folder. 
4. load the `vegan` R package (be sure to install if needed).  

```{r}
getwd()
setwd("C:/Users/ADMIN/OneDrive/Documents/GitHub/QB2025_Nambiar/Week4-Beta")

library(vegan)
library(ade4)
library(indicspecies)

```

## 2) LOADING DATA
### Load dataset
In the R code chunk below, load the `doubs` dataset from the `ade4` package

```{r}
# note, pleae do not print the dataset when submitting

data(doubs)


```
## 3) HYPOTHESIS TESTING

### A.  Multivariate Procedures for Categorical Designs
Earlier work done in the Doubs River suggested that the river has four distinct regions of habitat quality: the first region (sites 1-14) of "high quality"; the second (sites 15 - 19) and fourth (sites 26 - 30) of "moderate quality"; and the third  (sites 20 - 25) of "low quality".

In the code chunk below, test the hypothesis that fish community composition varies with river quality.

1. create a factor vector that categorizes habitat quality in the Doubs River,
2. use the multivariate analyses for categorical predictors to describe how fish community structure relates to habitat quality.

```{r}
data(doubs)

# factors vector
fish<-doubs$fish
fish<-fish[-8,]
dim(fish)
quality<-c(rep("HQ",13), rep("MQ",5), rep("LQ", 6), rep("MQ", "5"))
length(quality)
#run PERMANOVA with adonis function

adonis2(fish~quality, method="bray", permutation=999) #Permanova

indval<-multipatt(fish, cluster=quality, func="IndVal.g", control=how(nperm=999))
summary(indval)


fish.rel<-decostand(fish,method="total")
phi<-multipatt(fish.rel,cluster=quality, func="r.g", control=how(nperm=999))

summary(phi)

```

***Question 1***: Based on the PERMANOVA, IndVal, and phi coefficient analyses, what did you learn about the relationship between habitat quality and the fish species composition?
Are the different analyses consistent with one another and do they agree with the visualizations (heat maps, cluster dendograms, ordinations) that you created?

> ***Answer 1***: 
Through PERMANOVA we can tell that there are significant differences in habitat types based on species composition, the INDval analysis help tell us about "indicator species" that are associated with certain habitat characteristics and we see that Alal, Gogo and Ruru are some of the strongest indicators. The phi coefficients help with the quantification of species habitat - association - biggest takwawy here is that 9 species are associated with a single habitat type while 9 are associated with 2 habitats.  Our results hear are consistent with the visialuzations that was created in the previous assignment. 

### B.  Multivariate Procedures for Continuous Designs

### i.  Mantel Test
In the R code chunk below, do the following:   

1. create distance matrices for both fish communities and environmental factors, and
2. use a Mantel test to test the hypothesis that fish assemblages are correlated with stream environmental variables.

```{r}
#data(doubs)
fish.dist<-vegdist(fish, method="bray")
env.dist<-vegdist(scale(doubs$env[-8,]),method="euclid")

#mantel test
mantel(fish.dist,env.dist)

```

***Question 2***: What do the results from our Mantel test suggest about fish diversity and stream environmental conditions? 
How does this relate to your hypothesis about stream quality influencing fish communities?

> ***Answer 2***:  Based off the significant mantel's correlation beteen community composition and the habitat enviornmental variables i.e. what is the in  doubs$en we see that environmental conditions are linked with species composition. Hence similarity in env conditions would liekly lead to similarity in community composition.  

### ii.  Constrained Ordination
In the R code chunk below, do the following:  

1. create an environmental matrix of the water chemistry data included in the `doubs` dataset using forward and reverse selection of variables,
2. conduct a redundancy analysis on the fish assemblages of the Doubs River,
3. use a permutation test to determine the significance of the constrained analysis,
4. use a permutation test to determine the correlation of each environmental factor on the constrained axes,
5. calculate the explained variation on the first and second constrained axes,
6. plot the constrained ordination results including labeled points for each site, and
7. add vectors that demonstrate the influence of each environmental factor the constrained ordination.   
```{r}
#env variables

env.chem<-as.matrix(doubs$env[-8,5:11])

fish.db<-vegdist(fish, method="bray", upper=TRUE, diag = T)

#perform dbRDA

doubs.dbrda<-dbrda(fish.db~.,as.data.frame(env.chem))

ordiplot(doubs.dbrda) #ordination plot

# Null model (intercept only)
doubs.dbrda.mod0 <- dbrda(fish.db ~ 1, as.data.frame(env.chem))
ordiplot(doubs.dbrda.mod0) # Simple MDS (i.e., PCoA)

# Full model (all explanatory variables)
doubs.dbrda.mod1 <- dbrda(fish.db ~ ., as.data.frame(env.chem))

# Stepwise selection (AIC-based)
doubs.dbrda <- ordiR2step(doubs.dbrda.mod0, doubs.dbrda.mod1, perm.max = 200)

# View selected model
doubs.dbrda$call
doubs.dbrda$anova
ordiplot(doubs.dbrda)

# Permutation tests for significance
permutest(doubs.dbrda, permutations = 999)
envfit(doubs.dbrda, env.chem[, c(4,6,7)], perm = 999)

# Explained variation
dbrda.explainvar1 <- round(doubs.dbrda$CCA$eig[1] / 
  sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100
dbrda.explainvar2 <- round(doubs.dbrda$CCA$eig[2] / 
  sum(c(doubs.dbrda$CCA$eig, doubs.dbrda$CA$eig)), 3) * 100


# Set plot margins
par(mar = c(5, 5, 4, 4) + 0.1)

# Initialize plot
plot(scores(doubs.dbrda, display = "wa"), xlim = c(-1.3, 1.1), ylim = c(-1.1, 2.7),
     xlab = paste("dbRDA 1 (", dbrda.explainvar1, "%)", sep = ""),
     ylab = paste("dbRDA 2 (", dbrda.explainvar2, "%)", sep = ""),
     pch = 16, cex = 2.0, type = "n", cex.lab = 1.5, cex.axis = 1.2, axes = FALSE)

# Add axes
axis(side = 1, labels = TRUE, lwd.ticks = 2, cex.axis = 1.2, las = 1)
axis(side = 2, labels = TRUE, lwd.ticks = 2, cex.axis = 1.2, las = 1)
abline(h = 0, v = 0, lty = 3)
box(lwd = 2)

# Add points and labels
points(scores(doubs.dbrda, display = "wa"), pch = 19, cex = 3, bg = "gray", col = "gray")
text(scores(doubs.dbrda, display = "wa"), labels = row.names(scores(doubs.dbrda, display = "wa")))

# Add environmental vectors
vectors <- scores(doubs.dbrda, display = "bp")
arrows(0, 0, vectors[,1], vectors[,2], lwd = 2, lty = 1, length = 0.2, col = "red")
text(vectors[,1], vectors[,2], pos = 3, labels = row.names(vectors))

# Add additional axes for vectors
axis(side = 3, lwd.ticks = 2, cex.axis = 1.2, las = 1, col = "red", lwd = 2.2,
     at = pretty(range(vectors[,1])), labels = pretty(range(vectors[,1])))
axis(side = 4, lwd.ticks = 2, cex.axis = 1.2, las = 1, col = "red", lwd = 2.2,
     at = pretty(range(vectors[,2])) * 2, labels = pretty(range(vectors[,2])))










```

***Question 3***: Based on the constrained ordination, what are the environmental variables (or groups of correlated variables) that seem to be contributing to variation in fish community structure?

> ***Answer 3***:  We see that the largest contributors to the variation in the env variables are bdo, oxy and nit.Wee see this from the dbRDA analysis i.e. all variables with high rsquared values.The DbRDA plot can be used to see the most significant variables and also covariation between the environmental variables.  

### iii. Variation Partitioning
In the code chunk below,   

1. Create a matrix model of the selected environmental variables,
2. Create a matrix model of the selected PCNM axes,
3. Perform constrained and partial constrained ordinations using the spatial and environmental models you just created,
4. Test the significance of each of your constrained ordinations using permutation tests,
5. Partition the variation among sites into the relative importance of space, environment, spatially structured environment, and residuals,
6. Plot the variation partitioning output to visualize it.  
```{r}

doubs.dbrda$anova

env.mod<-model.matrix(~oxy+bdo+nit,as.data.frame(env.chem))[,-1]

fish<-doubs$fish[-8, ]
dim(fish)
# Weight each site by relative abundance
rs <- rowSums(fish) / sum(fish)
xy<-doubs$xy[-8, ]
dim(xy)
# Perform PCNM
doubs.pcnm <- pcnm(dist(xy), w = rs, dist.ret = TRUE)

# Keep only meaningful eigenvectors (positive eigenvalues)

# Find valid eigenvectors with positive eigenvalues
valid_eigenvectors <- which(doubs.pcnm$values[1:ncol(doubs.pcnm$vectors)] > 0)

# Check if valid eigenvectors exist
if (length(valid_eigenvectors) > 0) {
    doubs.pcnm$vectors <- doubs.pcnm$vectors[, valid_eigenvectors, drop = FALSE]
} else {
    stop("No positive eigenvalues found. Check PCNM computation.")
}



# Model selection for PCNM axes
doubs.space <- as.data.frame(scores(doubs.pcnm))
doubs.pcnm.mod0 <- dbrda(fish.db ~ 1, doubs.space)
doubs.pcnm.mod1 <- dbrda(fish.db ~ ., doubs.space)
step.pcnm <- ordiR2step(doubs.pcnm.mod0, doubs.pcnm.mod1, perm.max = 200)

# View selected model
step.pcnm$anova

# Construct spatial model using selected PCNM axes
space.mod <- model.matrix(~ PCNM2 + PCNM3 + PCNM5 + PCNM1 + PCNM13 + PCNM16 + PCNM6, doubs.space)[, -1]

# Constrained ordinations
doubs.total.env <- dbrda(fish.db ~ env.mod)
doubs.total.space <- dbrda(fish.db ~ space.mod)

# Partial constrained ordinations
doubs.env.cond.space <- dbrda(fish.db ~ env.mod + Condition(space.mod))
doubs.space.cond.env <- dbrda(fish.db ~ space.mod + Condition(env.mod))

# Test significance of dbRDA fractions
permutest(doubs.env.cond.space, permutations = 999)
permutest(doubs.space.cond.env, permutations = 999)
permutest(doubs.total.env, permutations = 999)
permutest(doubs.total.space, permutations = 999)

# Perform variation partitioning
doubs.varpart <- varpart(fish.db, env.mod, space.mod)
summary(doubs.varpart)

# Set plot margins and plot variation partitioning
plot.new()
par(mar = c(2,2,2,2))
plot(doubs.varpart)

#unable to run the text code. It says plot.new has not been called yet, even when I run plot.new()

```

***Question 4***: Interpret the variation partitioning results.   

> ***Answer 4***: Overall, we see that there are both environmental and spatial effects along with a combined effect of both. Based off the summary of the varpart we see that the water quality (varaibles together) controbutes to 13% of the total variation, Spatial effects explain 23% of the total variation, a combination of both effects (since resources depend on spatial factors) explains 18% of the variation. All of this accoutn for 55% of what explains fish community composition while the rest - 45% is the residual variation that is not account for by any of the variables in this dataset. 

## SYNTHESIS

Load the dataset from that you and your partner are using for the team project. 
Use one of the hypothesis-testing tools introduced in the beta diversity module. 
Interpret the findings of your data with respect to principles of biodiversity. 

